---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

The code below loads the activity data into R and also converts the date into the correct format.

```{r echo = TRUE}
data <- read.csv("activity.csv", colClasses = c("integer", "Date", "integer"))
```

The packages used in the analysis also need to be loaded.

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```

## What is the mean total number of steps taken per day?

The total number of steps taken per day is calculated using the below code. A summary of the table is provided for reference.

```{r echo = TRUE, message = FALSE}
totalstep <- data %>% group_by(date) %>% summarise(totalsteps = sum(steps))
totalstep
```

The following histogram shows the total number of steps taken per day using the base plotting system.

```{r echo = TRUE}
hist(totalstep$totalsteps, main = "Total Number of Steps per Day", xlab = "Total Steps")
```

The mean and median of the total number of steps is reported below, with the missing values being removed from the calculation.

```{r echo = TRUE}
print(mean(totalstep$totalsteps, na.rm = TRUE))
print(median(totalstep$totalsteps, na.rm = TRUE))
```

## What is the average daily activity pattern?

The following time series plot shows the average number of steps taken (averaged across all days) by 5-minute intervals. The base plotting system has been used for this graph.

```{r echo = TRUE, message = FALSE}
avgint <- data %>% group_by(interval) %>% summarise(avgint = mean(steps, na.rm = TRUE))

with(avgint, plot(interval, avgint, type = "l", xlab = "5-minute Interval", ylab = "Average Number of Steps"))
with(avgint, title(main = "Average Number of Steps Per Day by 5-minute Interval"))
```

The 5-minute interval, on average across all days, containing the maximum number of steps is 835, as reported below:

```{r echo = TRUE}
print(avgint[which.max(avgint$avgint), ])
```

## Imputing missing values

The code below shows the total number of missing values in the dataset.

```{r echo = TRUE}
print(sum(is.na(data)))
```

As noted above, there are a number of missing values in the dataset which require filling in. As the time of day makes a substantial difference in the number of steps taken, I have used the mean for the relevant 5-minute interval to replace the missing values. 

The following code creates a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo = TRUE}
newdata <- data
newdata$steps <- ave(newdata$steps, newdata$int, FUN = function(x)
    ifelse(is.na(x), mean(x, na.rm = TRUE), x))
```

The following histogram (created using the base plotting system) shows the total number of steps taken per day with the missing data imputed.

```{r echo = TRUE, message = FALSE}
totalstep2 <- newdata %>% group_by(date) %>% summarise(totalsteps = sum(steps))
hist(totalstep2$totalsteps, main = "Total Number of Steps per Day", xlab = "Total Steps")
```

Below is the mean and median of the total number of steps taken per day.

```{r echo = TRUE}
print(mean(totalstep2$totalsteps, na.rm = TRUE))
print(median(totalstep2$totalsteps, na.rm = TRUE))
```

The mean has is consistent between the original and the new data set, whereas the median has increased by a minor amount. This change is due to the fact that the missing data was imputed using the mean of the relevant 5-minute interval. 

By imputing missing data, total number of steps increases as steps are being added into the dataset based on the mean of non-missing values. 

## Are there any differences in activity patterns between weekdays and weekends?

The code below creates a new factor varaible with two levels ("weekday" and "weekend") which indicates whether a given date is a weekday or a weekend day. 

```{r echo = TRUE}
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
newdata$DateClass <- factor((weekdays(newdata$date) %in% weekday), 
    levels = c(FALSE, TRUE), labels = c("weekend", "weekday"))
```

The following panel plot shows time series data of the 5-minute interval and the average number of steps taken, averaged across weekdays and weekends. This plot was created using the ggplot2 package.

```{r echo = TRUE}
avgweekclass <- aggregate(steps ~ DateClass + interval, data = newdata, mean)

ggplot(avgweekclass, aes(interval, steps)) + geom_line() + facet_grid(DateClass~.) +
    labs(title = "Average Number of Steps by Weekday/Weekend per 5-minute Interval",
         x = "Interval", y = "Number of Steps")
```